---
title: "Motivating example for application of SNN to single cell cancer data"
author: "Yusha Liu"
date: "2022-1-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Summary
We present a toy example which illustrates the characteristics observed in single cell cancer data, which demonstrate strong inter-patient heterogeneity but still share similarities among patients.

## Simulation setup
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(ebnm)
library(flashier)
library(pheatmap)
library(gridExtra)
library(tidyverse)

### load in simulated loadings and factors
L <- readRDS("data/model.rds")$L
F <- readRDS("data/model.rds")$F
F[,1] <- F[,1] - 10

### plot the poplulation factorization of truth, where F is orthogonal across columns 
L.pop <- L[, 1:9]
F.pop <- F[, 1:9]
F.pop[,2:5] <- F[,2:5] + F[,10]
F.pop[,6:9] <- F[,6:9] + F[,11]
plt.L <- pheatmap(L.pop[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
                  main = "True L\n population factorization", silent = TRUE)
plt.F <- pheatmap(F.pop[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
                  main = "True F\n population factorization", silent = TRUE)
grid.arrange(plt.L$gtable, plt.F$gtable, ncol=2)

### plot the drift factorization of truth, where F is orthogonal across columns 
plt.L <- pheatmap(L[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main = "True L\n drift factorization", silent = TRUE)
plt.F <- pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main = "True F\n drift factorization", silent = TRUE)
grid.arrange(plt.L$gtable, plt.F$gtable, ncol=2)

### simulate the data
set.seed(100)
X <- L %*% t(F) + matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)
XXt <- X %*% t(X)/ncol(X)

```


### Define functions to fit SNN
```{r warning=FALSE, message=FALSE}
source("code/functions.R")
```


### Define plot functions
```{r warning=FALSE, message=FALSE}

# plot posterior mean of L and F after rescaling
plotLF <- function(fit, title){
  fit.L <- fit$loadings.pm[[1]][, fit$pve > 0]
  fit.L <- t(t(fit.L)/apply(fit.L, 2, max))
  colnames(fit.L) <- paste0("k", 1:ncol(fit.L))
  fit.F <- fit$loadings.pm[[2]][, fit$pve > 0]
  fit.F <- t(t(fit.F)/apply(fit.F, 2, max))
  colnames(fit.F) <- paste0("k", 1:ncol(fit.F))
  plt.L <- pheatmap(fit.L, cluster_rows = FALSE, cluster_cols = FALSE, silent = TRUE, main=title[1])
  plt.F <- pheatmap(fit.F, cluster_rows = FALSE, cluster_cols = FALSE, silent = TRUE, main=title[2])
  grid.arrange(plt.L$gtable, plt.F$gtable, ncol=2)
}

```


### Fit SNN to the data matrix
As can be seen, SNN fit to the cell $\times$ gene data matrix (point-exponential prior on loadings, point-Laplace prior on factors) yields a population factorization, in which the loadings fail to reveal the similarities among patients.  
```{r warning=F}

### initialize by running flash without constraints on L
fit.pn <- flash.init(X, S = 1/sqrt(nrow(X)), var.type = 2) %>% flash.add.greedy(Kmax = 11, prior.family = prior.point.laplace()) %>% flash.backfit()
snn.init <- init.snn.LL(fit.pn)

### run SNN with point exponential prior on L
fit.snn <- flash.init(X, S = 1/sqrt(nrow(X)), var.type = 2
) %>% flash.init.factors(EF = snn.init, prior.family = c(as.prior(ebnm::ebnm_point_exponential, sign = 1), prior.point.laplace())
) %>% flash.backfit()

### plot the estimated L
fit.L <- fit.snn$loadings.pm[[1]][, fit.snn$pve > 0]
fit.L <- t(t(fit.L)/apply(fit.L, 2, max))
colnames(fit.L) <- paste0("k", 1:ncol(fit.L))
pheatmap(fit.L, cluster_rows = FALSE, cluster_cols = FALSE, main="L from SNN fit to data matrix")

```


### Fit SNN to the covariance matrix with a point-exponential prior
To encourage a drift factorization in which the loadings reveal similarities among patients and are thus more informative, we fit SNN to the cell $\times$ cell covariance matrix to estimate the loadings first, and then fit SNN to the cell $\times$ gene data matrix by keeping loadings fixed.   
```{r warning=F}

### fit unconstrained flash with point Laplace prior to XXt while ignoring the diagonal component for now
fit.cov <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 11, prior.family = prior.point.laplace()) %>% flash.backfit()

### fit unconstrained flash again with the diagonal component
fit.cov <- fit.ebcovmf(dat=XXt, fl=fit.cov, prior=prior.point.laplace())$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit.cov)

### fit flash with point exponential prior to XXt while ignoring the diagonal component for now
fit.cov <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)) 

### fit flash again with the diagonal component
fit.cov <- fit.ebcovmf(dat=XXt, fl=fit.cov, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### plot the estimated L and F
plotLF(fit.cov, title=c("L from SNN fit to covariance\n point exponential prior)", "F from SNN fit to covariance\n point exponential prior)"))

```


### Fit SNN to the covariance matrix with a generalized binary prior 
In the SNN fit to the covariance matrix with a point-exponential prior, the loadings do reveal similaries among patients, but the solution still looks quite noisy and is not close to the drift factorization desired. 

To fix this, we instead fit SNN to the covariance matrix with a generalized binary prior, which can be expressed as
\begin{align}
g \sim (1-\pi) \delta_0 + \pi N_{[0, \infty)}(\mu, \sigma^2),
\end{align}
where the truncated normal component is highly spiky, i.e., $\frac{\mu}{\sigma}$ is large.

```{r warning=F}

### fit flash with generalized binary prior to XXt while ignoring the diagonal component for now
fit.cov <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm_binary_general, sign = 1)) 

### fit flash again with the diagonal component
fit.cov <- fit.ebcovmf(dat=XXt, fl=fit.cov, prior=as.prior(ebnm_binary_general, sign = 1))$fl

### plot the estimated L and F
plotLF(fit.cov, title=c("L from SNN fit to covariance\n general binary prior)", "F from SNN fit to covariance\n general binary prior)"))

```