---
title: "Application of flash to covariance matrix with generalized binary prior"
author: "Yusha Liu"
date: "2022-3-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Summary
We fit flash to the covariance matrix with generalized binary prior on each loading/factor, i.e, $l_k \sim g_{lk}$ and $f_k \sim g_{fk}$ such that
\begin{align}
g \sim (1-\pi) N(\mu_1, \sigma^2) + \pi N(\mu_2, \sigma^2),
\end{align}
where $\frac{\mu_2 - \mu_1}{\sigma}$ is large. 

A further generalization is to allow difference variances for each normal mixture component, such that
\begin{align}
g \sim (1-\pi) N(\mu_1, \sigma_1^2) + \pi N(\mu_2, \sigma_2^2),
\end{align}
where $\frac{\mu_2 - \mu_1}{\sigma_1}$ and $\frac{\mu_2 - \mu_1}{\sigma_2}$ are both large. 


## Simulation setup
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(flashier)
library(pheatmap)
library(gridExtra)
library(tidyverse)

### load in simulated loadings and factors
L <- readRDS("data/model.rds")$L
F <- readRDS("data/model.rds")$F
F[,1] <- F[,1] - 10

### plot the drift factorization of truth, where F is orthogonal across columns 
plt.L <- pheatmap(L[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main = "True L", silent = TRUE)
plt.F <- pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main = "True F", silent = TRUE)
grid.arrange(plt.L$gtable, plt.F$gtable, ncol=2)

```


### Define plot functions
```{r warning=FALSE, message=FALSE}

### define the function to plot scaled L and F
plotLF_scaled <- function(fit.cov, title){
  k.order <- order(fit.cov$pve, decreasing = TRUE)
  fit.L <- fit.cov$L.pm[, k.order]
  fit.F <- fit.cov$F.pm[, k.order]
  fit.L <- fit.L[, apply(fit.L, 2, sd)!=0]
  fit.F <- fit.F[, apply(fit.F, 2, sd)!=0]
  new.L <- fit.L[, -1]
  new.F <- fit.F[, -1]
  colnames(new.L) <- paste0("k", 2:ncol(fit.L))
  colnames(new.F) <- paste0("k", 2:ncol(fit.F))
  
  for(k in 2:ncol(fit.L)){
    k.cur <- k.order[k]
    new.L[, k-1] <- (fit.L[,k] - fit.cov$L.ghat[[k.cur]]$mean[1])/(fit.cov$L.ghat[[k.cur]]$mean[2] - fit.cov$L.ghat[[k.cur]]$mean[1])
    new.F[, k-1] <- (fit.F[,k] - fit.cov$F.ghat[[k.cur]]$mean[1])/(fit.cov$F.ghat[[k.cur]]$mean[2] - fit.cov$F.ghat[[k.cur]]$mean[1])
  }
  
  plt.L <- pheatmap(new.L, cluster_rows = FALSE, cluster_cols = FALSE, silent = TRUE, main = title[1])
  plt.F <- pheatmap(new.F, cluster_rows = FALSE, cluster_cols = FALSE, silent = TRUE, main = title[2])
  grid.arrange(plt.L$gtable, plt.F$gtable, ncol=2)
}

```


## Simulation results
### rep=1
```{r warning=F}

### fit flash with general binary prior by initialization using truth followed by backfit
fit1 <- readRDS("output/general_binary_prior_rep1_init1.Rds")

### fit flash with general binary prior with the same variance, through greedy + backfit
fit2 <- readRDS("output/general_binary_prior_rep1_init2.Rds")

### fit flash with general binary prior with different variances, through greedy + backfit
fit3 <- readRDS("output/general_binary_prior_rep1_init3.Rds")

### fit flash with general binary prior with the same variance, by initialization using point Laplace + backfit
fit4 <- readRDS("output/general_binary_prior_rep1_init4.Rds")

### fit flash with general binary prior with different variances, by initialization using point Laplace + backfit
fit5 <- readRDS("output/general_binary_prior_rep1_init5.Rds")

### compare the elbo
fit1$elbo
fit2$elbo
fit3$elbo
fit4$elbo
fit5$elbo

### plot the estimated L and F on the rescaled scale
plotLF_scaled(fit1, title=c("rep1: rescaled L (equal variance)\n oracle init", "rep1: rescaled F (equal variance)\n oracle init"))
plotLF_scaled(fit2, title=c("rep1: rescaled L (equal variance)\n greedy + backfit", "rep1: rescaled F (equal variance)\n greedy + backfit"))
plotLF_scaled(fit3, title=c("rep1: rescaled L (unequal variance)\n greedy + backfit", "rep1: rescaled F (unequal variance)\n greedy + backfit"))
plotLF_scaled(fit4, title=c("rep1: rescaled L (equal variance)\n point Laplace init", "rep1: rescaled F (equal variance)\n point Laplace init"))
plotLF_scaled(fit5, title=c("rep1: rescaled L (unequal variance)\n point Laplace init", "rep1: rescaled F (unequal variance)\n point Laplace init"))

```


### rep=2
```{r warning=F}

### fit flash with general binary prior by initialization using truth followed by backfit
fit1 <- readRDS("output/general_binary_prior_rep2_init1.Rds")

### fit flash with general binary prior with the same variance, through greedy + backfit
fit2 <- readRDS("output/general_binary_prior_rep2_init2.Rds")

### fit flash with general binary prior with different variances, through greedy + backfit
fit3 <- readRDS("output/general_binary_prior_rep2_init3.Rds")

### fit flash with general binary prior with the same variance, by initialization using point Laplace + backfit
fit4 <- readRDS("output/general_binary_prior_rep2_init4.Rds")

### fit flash with general binary prior with different variances, by initialization using point Laplace + backfit
fit5 <- readRDS("output/general_binary_prior_rep2_init5.Rds")

### compare the elbo
fit1$elbo
fit2$elbo
fit3$elbo
fit4$elbo
fit5$elbo

### plot the estimated L and F on the rescaled scale
plotLF_scaled(fit1, title=c("rep2: rescaled L (equal variance)\n oracle init", "rep2: rescaled F (equal variance)\n oracle init"))
plotLF_scaled(fit2, title=c("rep2: rescaled L (equal variance)\n greedy + backfit", "rep2: rescaled F (equal variance)\n greedy + backfit"))
plotLF_scaled(fit3, title=c("rep2: rescaled L (unequal variance)\n greedy + backfit", "rep2: rescaled F (unequal variance)\n greedy + backfit"))
plotLF_scaled(fit4, title=c("rep2: rescaled L (equal variance)\n point Laplace init", "rep2: rescaled F (equal variance)\n point Laplace init"))
plotLF_scaled(fit5, title=c("rep2: rescaled L (unequal variance)\n point Laplace init", "rep2: rescaled F (unequal variance)\n point Laplace init"))

```


### rep=3
```{r warning=F}

### fit flash with general binary prior by initialization using truth followed by backfit
fit1 <- readRDS("output/general_binary_prior_rep3_init1.Rds")

### fit flash with general binary prior with the same variance, through greedy + backfit
fit2 <- readRDS("output/general_binary_prior_rep3_init2.Rds")

### fit flash with general binary prior with different variances, through greedy + backfit
fit3 <- readRDS("output/general_binary_prior_rep3_init3.Rds")

### fit flash with general binary prior with the same variance, by initialization using point Laplace + backfit
fit4 <- readRDS("output/general_binary_prior_rep3_init4.Rds")

### fit flash with general binary prior with different variances, by initialization using point Laplace + backfit
fit5 <- readRDS("output/general_binary_prior_rep3_init5.Rds")

### compare the elbo
fit1$elbo
fit2$elbo
fit3$elbo
fit4$elbo
fit5$elbo

### plot the estimated L and F on the rescaled scale
plotLF_scaled(fit1, title=c("rep3: rescaled L (equal variance)\n oracle init", "rep3: rescaled F (equal variance)\n oracle init"))
plotLF_scaled(fit2, title=c("rep3: rescaled L (equal variance)\n greedy + backfit", "rep3: rescaled F (equal variance)\n greedy + backfit"))
plotLF_scaled(fit3, title=c("rep3: rescaled L (unequal variance)\n greedy + backfit", "rep3: rescaled F (unequal variance)\n greedy + backfit"))
plotLF_scaled(fit4, title=c("rep3: rescaled L (equal variance)\n point Laplace init", "rep3: rescaled F (equal variance)\n point Laplace init"))
plotLF_scaled(fit5, title=c("rep3: rescaled L (unequal variance)\n point Laplace init", "rep3: rescaled F (unequal variance)\n point Laplace init"))

```


### rep=4
```{r warning=F}

### fit flash with general binary prior by initialization using truth followed by backfit
fit1 <- readRDS("output/general_binary_prior_rep4_init1.Rds")

### fit flash with general binary prior with the same variance, through greedy + backfit
fit2 <- readRDS("output/general_binary_prior_rep4_init2.Rds")

### fit flash with general binary prior with different variances, through greedy + backfit
fit3 <- readRDS("output/general_binary_prior_rep4_init3.Rds")

### fit flash with general binary prior with the same variance, by initialization using point Laplace + backfit
fit4 <- readRDS("output/general_binary_prior_rep4_init4.Rds")

### fit flash with general binary prior with different variances, by initialization using point Laplace + backfit
fit5 <- readRDS("output/general_binary_prior_rep4_init5.Rds")

### compare the elbo
fit1$elbo
fit2$elbo
fit3$elbo
fit4$elbo
fit5$elbo

### plot the estimated L and F on the rescaled scale
plotLF_scaled(fit1, title=c("rep4: rescaled L (equal variance)\n oracle init", "rep4: rescaled F (equal variance)\n oracle init"))
plotLF_scaled(fit2, title=c("rep4: rescaled L (equal variance)\n greedy + backfit", "rep4: rescaled F (equal variance)\n greedy + backfit"))
plotLF_scaled(fit3, title=c("rep4: rescaled L (unequal variance)\n greedy + backfit", "rep4: rescaled F (unequal variance)\n greedy + backfit"))
plotLF_scaled(fit4, title=c("rep4: rescaled L (equal variance)\n point Laplace init", "rep4: rescaled F (equal variance)\n point Laplace init"))
plotLF_scaled(fit5, title=c("rep4: rescaled L (unequal variance)\n point Laplace init", "rep4: rescaled F (unequal variance)\n point Laplace init"))

```


### rep=5
```{r warning=F}

### fit flash with general binary prior by initialization using truth followed by backfit
fit1 <- readRDS("output/general_binary_prior_rep5_init1.Rds")

### fit flash with general binary prior with the same variance, through greedy + backfit
fit2 <- readRDS("output/general_binary_prior_rep5_init2.Rds")

### fit flash with general binary prior with different variances, through greedy + backfit
fit3 <- readRDS("output/general_binary_prior_rep5_init3.Rds")

### fit flash with general binary prior with the same variance, by initialization using point Laplace + backfit
fit4 <- readRDS("output/general_binary_prior_rep5_init4.Rds")

### fit flash with general binary prior with different variances, by initialization using point Laplace + backfit
fit5 <- readRDS("output/general_binary_prior_rep5_init5.Rds")

### compare the elbo
fit1$elbo
fit2$elbo
fit3$elbo
fit4$elbo
fit5$elbo

### plot the estimated L and F on the rescaled scale
plotLF_scaled(fit1, title=c("rep5: rescaled L (equal variance)\n oracle init", "rep5: rescaled F (equal variance)\n oracle init"))
plotLF_scaled(fit2, title=c("rep5: rescaled L (equal variance)\n greedy + backfit", "rep5: rescaled F (equal variance)\n greedy + backfit"))
plotLF_scaled(fit3, title=c("rep5: rescaled L (unequal variance)\n greedy + backfit", "rep5: rescaled F (unequal variance)\n greedy + backfit"))
plotLF_scaled(fit4, title=c("rep5: rescaled L (equal variance)\n point Laplace init", "rep5: rescaled F (equal variance)\n point Laplace init"))
plotLF_scaled(fit5, title=c("rep5: rescaled L (unequal variance)\n point Laplace init", "rep5: rescaled F (unequal variance)\n point Laplace init"))

```