---
title: "Application of SNN to covariance matrix of simulated Gaussian data with correlated factors"
author: "Yusha Liu"
date: "2021-10-14"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation setup
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(flashier)
library(magrittr)
library(RColorBrewer)
library(pheatmap)

### load in the function to implement snn
source("code/util.R")

### set seed
set.seed(100)

### load in simulated loadings and factors
L <- readRDS("data/model.rds")$L
F <- readRDS("data/model.rds")$F

### plot the drift factorization of truth, where F is orthogonal across columns 
pheatmap(L[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True L")
pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")

```


### Use true L for initialization (sigma2=1)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)

### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit1 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=list(L,L), prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)

### fit flash again with the diagonal component
fit1 <- fit.ebcovmf(dat=XXt, fl=fit1, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit1.L <- fit1$loadings.pm[[1]]
fit1.L <- fit1.L[, fit1$pve > 0]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (oracle init, sigma2=1)")

```


### Add all factors greedily, then take nonnegative transform and backfit all factors (sigma2=1)
```{r warning=F}
### fit unconstrained flash with point Laplace prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 11, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit2)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)

### fit flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit2.L <- fit2$loadings.pm[[1]]
fit2.L <- fit2.L[, fit2$pve > 0]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L with point-exponential prior \n (greedy all + backfit all, sigma2=1")

```


### Iteratively add each factor greedily, then take nonnegative transform and backfit that factor (sigma2=1)
```{r warning=F}
### fit flash with point exponential prior to XXt 
fit3 <- fit.ebcovmf.snn(dat=XXt, Kmax=11)

### look at the estimates of L
fit3.L <- fit3$loadings.pm[[1]]
fit3.L <- fit3.L[, fit3$pve > 0]
fit3.L <- t(t(fit3.L)/apply(fit3.L, 2, max))
colnames(fit3.L) <- paste0("k", 1:ncol(fit3.L))
pheatmap(fit3.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (iteratively greedy + backfit, sigma2=1)")

```


### Compare the three fits to the covariance matrix
```{r warning=F}
### number of factors
fit1$n.factors
fit2$n.factors
fit3$n.factors

### elbo
fit1$elbo
fit2$elbo
fit3$elbo

```


### Use true L for initialization (sigma2=0.5)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(800*2000, 0, 0.7), nrow=800, ncol=2000)

### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit1 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=list(L,L), prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)

### fit flash again with the diagonal component
fit1 <- fit.ebcovmf(dat=XXt, fl=fit1, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit1.L <- fit1$loadings.pm[[1]]
fit1.L <- fit1.L[, fit1$pve > 0]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (oracle init, sigma2=0.5)")

```


### Add all factors greedily, then take nonnegative transform and backfit all factors (sigma2=0.5)
```{r warning=F}
### fit unconstrained flash with point Laplace prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 11, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit2)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)

### fit flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit2.L <- fit2$loadings.pm[[1]]
fit2.L <- fit2.L[, fit2$pve > 0]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L with point-exponential prior \n (greedy all + backfit all, sigma2=0.5")

```


### Iteratively add each factor greedily, then take nonnegative transform and backfit that factor (sigma2=0.5)
```{r warning=F}
### fit flash with point exponential prior to XXt 
fit3 <- fit.ebcovmf.snn(dat=XXt, Kmax=11)

### look at the estimates of L
fit3.L <- fit3$loadings.pm[[1]]
fit3.L <- fit3.L[, fit3$pve > 0]
fit3.L <- t(t(fit3.L)/apply(fit3.L, 2, max))
colnames(fit3.L) <- paste0("k", 1:ncol(fit3.L))
pheatmap(fit3.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (iteratively greedy + backfit, sigma2=0.5)")

```


### Compare the three fits to the covariance matrix
```{r warning=F}
### number of factors
fit1$n.factors
fit2$n.factors
fit3$n.factors

### elbo
fit1$elbo
fit2$elbo
fit3$elbo

```