---
title: "Application of SNN to simulated Gaussian data with correlated factors"
author: "Yusha Liu"
date: "2021-10-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation setup
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(flashier)
library(magrittr)
library(RColorBrewer)
library(pheatmap)

### load in the function to implement snn
source("code/fit_snn.R")

### set seed
set.seed(100)

### load in simulated loadings and factors
L <- readRDS("data/model.rds")$L
F <- readRDS("data/model.rds")$F

### plot the drift factorization of truth, where F is orthogonal across columns 
pheatmap(L[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True L")
pheatmap(F[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")

```


### Fit SNN to X with point exponential prior on L using default initialization (sigma2=1)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(800*2000, 0, 1), nrow=800, ncol=2000)

### run snn on the simulated data
fit1.snn <- fit.snn(data=X, Kmax=12)

### define the color map for Z scores
cols <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
cols <- c(cols[1:25], "lightgray", cols[76:100])
brks <- c(seq(-6, -2, length=26), seq(2, 6, length=26))

### plot the scaled loadings and factors
fit1.L <- fit1.snn$loadings.pm[[1]]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
fit1.Z <- fit1.snn$loadings.pm[[2]]/fit1.snn$loadings.psd[[2]]
colnames(fit1.Z) <- paste0("k", 1:ncol(fit1.Z))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=1, default init)")
pheatmap(fit1.Z[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, 
         main="Estimated F (sigma2=1, default init)")

```


### Fit SNN to X with point exponential prior on L using true L and F for initialization (sigma2=1)
```{r warning=F}
### run snn on the simulated data
fit2.snn <- flash.init(X, S = 1/sqrt(nrow(X)), var.type = 2
) %>% flash.init.factors(EF = list(L, F), prior.family = c(as.prior(ebnm::ebnm_point_exponential, sign = 1), prior.normal.scale.mix())) %>% flash.backfit(verbose.lvl = 1)

### plot the scaled loadings and factors
fit2.L <- fit2.snn$loadings.pm[[1]]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
fit2.Z <- fit2.snn$loadings.pm[[2]]/fit2.snn$loadings.psd[[2]]
colnames(fit2.Z) <- paste0("k", 1:ncol(fit2.Z))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=1, oracle init)")
pheatmap(fit2.Z[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, 
         main="Estimated F (sigma2=1, oracle init)")

### compare the ELBO between two fits
fit1.snn$elbo
fit2.snn$elbo

```


### Fit flash to scaled XX' with point exponential prior on L (sigma2=1)
```{r warning=F}
### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### fit unconstrained flash with sparse prior to XXt without considering the diagonal component for now
fit.cov <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 12, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit.cov <- fit.ebcovmf(dat=XXt, fl=fit.cov)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit.cov)

### fit snn with point exponential prior to XXt without considering the diagonal component for now
fit.snn.cov <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)
kset <- fit.snn.cov$pve > 0
fit.snn.cov$flash.fit$EF <- lapply(fit.snn.cov$flash.fit$EF, function(x) x[, kset])
fit.snn.cov$flash.fit$EF2 <- lapply(fit.snn.cov$flash.fit$EF2, function(x) x[, kset])

### fit snn again with the diagonal component
fit.snn.cov <- fit.ebcovmf(dat=XXt, fl=fit.snn.cov, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit3.L <- fit.snn.cov$loadings.pm[[1]]
fit3.L <- fit3.L[, fit.snn.cov$pve > 0]
fit3.L <- t(t(fit3.L)/apply(fit3.L, 2, max))
colnames(fit3.L) <- paste0("k", 1:ncol(fit3.L))
pheatmap(fit3.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=1, fit to XXt)")

```


### Fit SNN to X with point exponential prior on L using default initialization (sigma2=0.5)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(800*2000, 0, 0.7), nrow=800, ncol=2000)

### run snn on the simulated data
fit1.snn <- fit.snn(data=X, Kmax=12)

### define the color map for Z scores
cols <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)
cols <- c(cols[1:25], "lightgray", cols[76:100])
brks <- c(seq(-6, -2, length=26), seq(2, 6, length=26))

### plot the scaled loadings and factors
fit1.L <- fit1.snn$loadings.pm[[1]]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
fit1.Z <- fit1.snn$loadings.pm[[2]]/fit1.snn$loadings.psd[[2]]
colnames(fit1.Z) <- paste0("k", 1:ncol(fit1.Z))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=0.5, default init)")
pheatmap(fit1.Z[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, 
         main="Estimated F (sigma2=0.5, default init)")

```


### Fit SNN to X with point exponential prior on L using true L and F for initialization (sigma2=0.5)
```{r warning=F}
### run snn on the simulated data
fit2.snn <- flash.init(X, S = 1/sqrt(nrow(X)), var.type = 2
) %>% flash.init.factors(EF = list(L, F), prior.family = c(as.prior(ebnm::ebnm_point_exponential, sign = 1), prior.normal.scale.mix())) %>% flash.backfit(verbose.lvl = 1)

### plot the scaled loadings and factors
fit2.L <- fit2.snn$loadings.pm[[1]]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
fit2.Z <- fit2.snn$loadings.pm[[2]]/fit2.snn$loadings.psd[[2]]
colnames(fit2.Z) <- paste0("k", 1:ncol(fit2.Z))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=0.5, oracle init)")
pheatmap(fit2.Z[,-1], show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = cols, breaks = brks, 
         main="Estimated F (sigma2=0.5, oracle init)")

### compare the ELBO between two fits
fit1.snn$elbo
fit2.snn$elbo

```


### Fit flash to scaled XX' with point exponential prior on L (sigma2=0.5)
```{r warning=F}
### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### fit unconstrained flash with sparse prior to XXt without considering the diagonal component for now
fit.cov <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 12, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit.cov <- fit.ebcovmf(dat=XXt, fl=fit.cov)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit.cov)

### fit snn with point exponential prior to XXt without considering the diagonal component for now
fit.snn.cov <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)
) %>% flash.backfit(verbose.lvl = 1)
kset <- fit.snn.cov$pve > 0
fit.snn.cov$flash.fit$EF <- lapply(fit.snn.cov$flash.fit$EF, function(x) x[, kset])
fit.snn.cov$flash.fit$EF2 <- lapply(fit.snn.cov$flash.fit$EF2, function(x) x[, kset])

### fit snn again with the diagonal component (got an error message when using a point exponential prior for L)
fit.snn.cov <- fit.ebcovmf(dat=XXt, fl=fit.snn.cov, prior=prior.nonnegative())$fl

### look at the estimates of L
fit3.L <- fit.snn.cov$loadings.pm[[1]]
fit3.L <- fit3.L[, fit.snn.cov$pve > 0]
fit3.L <- t(t(fit3.L)/apply(fit3.L, 2, max))
colnames(fit3.L) <- paste0("k", 1:ncol(fit3.L))
pheatmap(fit3.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="Estimated L (sigma2=0.5, fit to XXt)")

```