---
title: "Application of SNN to simulated Gaussian data with shared factors (a simple two-group example)"
author: "Yusha Liu"
date: "2021-10-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Simulation setup
```{r warning=FALSE, message=FALSE}
library(Matrix)
library(flashier)
library(magrittr)
library(RColorBrewer)
library(pheatmap)

### load in the function to implement snn
source("code/util.R")

### set seed
set.seed(1)

### simulate L and F
L <- matrix(0, nrow=200, ncol=3)
L[,1] <- 1
L[1:100, 2] <- 1
L[101:200, 3] <- 1
F <- matrix(0, nrow=1000, ncol=3)
F[1:100,1] <- rnorm(100, 0, 1)
F[101:500, 2] <- rnorm(400, 0, 1)
F[501:900, 3] <- rnorm(400, 0, 1)
colnames(L) <- paste0("k", 1:ncol(L))
colnames(F) <- paste0("k", 1:ncol(F))

### plot the drift factorization of truth, where F is orthogonal across columns 
pheatmap(L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True L")
pheatmap(F, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="True F")

```


### Factorize XXt: use true L for initialization (sigma2=1)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(200*1000, 0, 1), nrow=200, ncol=1000)

### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### plot the covariance matrix
pheatmap(XXt, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="XXt")

### calculate rescaled initial values for L
sf <- sqrt(nrow(F)/colSums(F^2))
init.F <- t(t(F)*sf)
init.L <- t(t(L)/sf)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit1 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=list(init.L, init.L), prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)) 

### fit flash again with the diagonal component
fit1 <- fit.ebcovmf(dat=XXt, fl=fit1, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit1.L <- fit1$loadings.pm[[1]]
fit1.L <- fit1.L[, fit1$pve > 0]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (oracle init, sigma2=1)")

```


### Factorize XXt: add all factors greedily, then take nonnegative transform and backfit all factors (sigma2=1)
```{r warning=F}
### fit unconstrained flash with point Laplace prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 3, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit2)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)) 

### fit flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit2.L <- fit2$loadings.pm[[1]]
fit2.L <- fit2.L[, fit2$pve > 0]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (greedy all + backfit all, sigma2=1)")

```


### Compare the fits to the covariance matrix
```{r warning=F}
### number of factors
sum(fit1$pve > 0)
sum(fit2$pve > 0)

### elbo
fit1$elbo
fit2$elbo

```


### Factorize XXt: use true L for initialization (sigma2=0.25)
```{r warning=F}
### simulate data with i.i.d. gaussian errors
X <- L %*% t(F) + matrix(rnorm(200*1000, 0, 0.5), nrow=200, ncol=1000)

### calculate the rescaled XX'
XXt <- X %*% t(X)/ncol(X)

### plot the covariance matrix
pheatmap(XXt, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, main="XXt")

### calculate rescaled initial values for L
sf <- sqrt(nrow(F)/colSums(F^2))
init.F <- t(t(F)*sf)
init.L <- t(t(L)/sf)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit1 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=list(init.L, init.L), prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)) 

### fit flash again with the diagonal component
fit1 <- fit.ebcovmf(dat=XXt, fl=fit1, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit1.L <- fit1$loadings.pm[[1]]
fit1.L <- fit1.L[, fit1$pve > 0]
fit1.L <- t(t(fit1.L)/apply(fit1.L, 2, max))
colnames(fit1.L) <- paste0("k", 1:ncol(fit1.L))
pheatmap(fit1.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (oracle init, sigma2=0.25)")

```


### Factorize XXt: add all factors greedily, then take nonnegative transform and backfit all factors (sigma2=0.25)
```{r warning=F}
### fit unconstrained flash with point Laplace prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.add.greedy(Kmax = 3, prior.family = prior.point.laplace(), verbose.lvl = 1
) %>% flash.backfit(verbose.lvl = 1)

### fit unconstrained flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2)$fl

### initialize the nonnegative fit based on the unconstrained flash fit 
snn.cov.init <- init.snn.cov(fit2)

### fit flash with point exponential prior to XXt without considering the diagonal component for now
fit2 <- flash.init(XXt, var.type = 0) %>% flash.init.factors(EF=snn.cov.init, prior.family=as.prior(ebnm::ebnm_point_exponential, sign = 1)) 

### fit flash again with the diagonal component
fit2 <- fit.ebcovmf(dat=XXt, fl=fit2, prior=as.prior(ebnm::ebnm_point_exponential, sign = 1))$fl

### look at the estimates of L
fit2.L <- fit2$loadings.pm[[1]]
fit2.L <- fit2.L[, fit2$pve > 0]
fit2.L <- t(t(fit2.L)/apply(fit2.L, 2, max))
colnames(fit2.L) <- paste0("k", 1:ncol(fit2.L))
pheatmap(fit2.L, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, 
         main="Estimated L with point-exponential prior \n (greedy all + backfit all, sigma2=0.25)")

```


### Compare the fits to the covariance matrix
```{r warning=F}
### number of factors
sum(fit1$pve > 0)
sum(fit2$pve > 0)

### elbo
fit1$elbo
fit2$elbo

```

